---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
For loading the data we assume that the *activity.zip* file is inside the working directory.  
We are going to load the libraries: dplyr, ggplot2.
```{r, echo=TRUE, results='hide'}
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
```
Also we will create a function to transform the time interval into time-formatted character.
```{r, echo=TRUE}
transf <- function(x) {paste(x %/% 100, x %% 100, sep = ":")}
```

Now loading the data and creating a new *timeinterval* column.
```{r, echo=TRUE}
myData <- read.table(unz("activity.zip","activity.csv"), header = TRUE, sep = ",", na.strings = "NA", colClasses = c("integer", "Date", "integer"))
myData$timeinterval <- transf(myData$interval)
```


## What is mean total number of steps taken per day?
1: Make a histogram of the total number of steps taken each day

```{r, echo=TRUE}
myDays <- group_by(myData, date)
myTotals <- summarise(myDays, steps = sum(steps, na.rm = TRUE))
hist(myTotals$steps, col = "green", xlab = "Total number of steps", main = "Histogram of total number of daily steps")
```

2: The mean is **`r mean(myTotals$steps)`** and the median is **`r median(myTotals$steps)`**.



## What is the average daily activity pattern?

1: Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}
myInterval <- group_by(myData, interval)
myAvg <- summarise(myInterval, steps = mean(steps, na.rm = TRUE))
plot(myAvg$interval, myAvg$steps, type = "l", xlab = "Interval", ylab = "Average number of steps", main = "Average number of steps per 5 minute interval", col = "red")
```

2: The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is `r transf(subset(myAvg, steps == max(myAvg$steps))[1,1])`.


## Imputing missing values

The total number of rows with NA values is `r sum(is.na(myData$steps))`.  
We will fill the NA values with average interval data. First, let's create the dataset with the average interval data column:

```{r, echo=TRUE}
myInterval <- group_by(myData, interval)
myInterval <- mutate(myInterval, average = mean(steps, na.rm = TRUE))
```

Then, create a new dataset *newData* equal to the original dataset but with the missing values filled in.
```{r, echo=TRUE}
newData <- myInterval
newData$steps[is.na(newData$steps)] <- newData$average[is.na(newData$steps)]
```

Let's make another histogram, this time with the new data:
```{r, echo=TRUE}
newDays <- group_by(newData, date)
newTotals <- summarise(newDays, steps = sum(steps))
hist(newTotals$steps, col = "green", xlab = "Total number of steps", main = "Histogram of total number of daily steps")


```

As we can see, the data is now more uniform than before.
Also, the new mean (`r as.integer(mean(newTotals$steps))`) and the new median (`r as.integer(median(newTotals$steps))`) of the total number of steps taken per day are different from the ones before since we have more data (NA values are replaced with positive values).



## Are there differences in activity patterns between weekdays and weekends?

In order to answer this question, there are a couple of steps involved.
1 - Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day. For this we create a function (wkdays) that we will use with sapply to create the variable.

```{r, echo=TRUE}
wkdays <- function(x) {
        val <- weekdays(x)
        if (val %in% c("Saturday", "Sunday")) {
                res <- "weekend"
        } else {
                res <- "weekday"
        }
        res
}
newData$wkdays <- as.factor(sapply(newData$date, wkdays))
```

2 - Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE}
myWkInterval <- group_by(newData, interval, wkdays)
myWkAvg <- summarize(myWkInterval, steps = mean(steps))

g <- ggplot(myWkAvg, aes(interval, steps))
g + geom_line(color = "steelblue") + facet_grid(wkdays ~ .)
```

So, yes, it seems there is a difference with weekdays peaking at around 9 AM while weekends have more steps across the day.